---
title: "K-Mean Clustering - Assignment 4"
author: "Shahbaz"
date: "2025-10-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Loading Libraries

```{r}
library(tidyverse)
library(cluster)
library(factoextra)
library(dplyr)
library(knitr)
```

# Loading Data

```{r}
pharma <- read.csv("/Users/shahbazshaikh/Downloads/Pharmaceuticals.csv")
summary(pharma)
str(pharma)
sum(is.na(pharma))
```

# Selecting only 9 Variables

```{r}

pharma_num <- pharma[, 3:11] 

pharma_scaled <- scale(pharma_num)
```

# Determining optimal number of K

```{r}
fviz_nbclust(pharma_scaled, kmeans, method = "wss") +
  geom_vline(xintercept = 3, linetype = 2) +
  labs(title = "Elbow Method for Optimal K - Elbow Point")
```

```{r}
fviz_nbclust(pharma_scaled, kmeans, method = "silhouette") +
  labs(title = "Silhouette Method for Optimal K")
```

# Performing K - Means Clustering

```{r}
set.seed(123)
optimal_k <- 3
k3 <- kmeans(pharma_scaled, centers = optimal_k, nstart = 25)
pharma$cluster <- as.factor(k3$cluster)
table(pharma$cluster)
```

# Visualizing the Clusters

```{r}
fviz_cluster(k3, data = pharma_scaled,
             ellipse.type = "norm",
             geom = "point",
             main = "K-means Clustering of Pharmaceutical Firms")
```

# Interpreting Cluster Characteristics

```{r}
cluster_summary <- pharma %>%
  mutate(Cluster = k3$cluster) %>%
  group_by(Cluster) %>%
  summarise(across(Market_Cap:Net_Profit_Margin, \(x) mean(x, na.rm = TRUE)))

cluster_summary
```

```{r}
pharma %>%
  mutate(Cluster = k3$cluster) %>%
  group_by(Cluster, Median_Recommendation) %>%
  summarise(count = n())

pharma %>%
  mutate(Cluster = k3$cluster) %>%
  group_by(Cluster, Location) %>%
  summarise(count = n())

pharma %>%
  mutate(Cluster = k3$cluster) %>%
  group_by(Cluster, Exchange) %>%
  summarise(count = n())
```

# Names for Cluster

```{r}
cluster_names <- c(
  "1" = "Established Global Leaders",
  "2" = "Balanced Performers", 
  "3" = "Emerging or Risky Firms"
)

# Apply cluster names to the dataset
pharma$cluster_named <- cluster_names[as.character(pharma$cluster)]

kable(pharma[, c("Name", "cluster", "cluster_named")])
```

#In this project, I used the Elbow Method to determine the optimal number of clusters. It helps identify the point where adding more clusters no longer significantly reduces the within-cluster variance. I chose this method because it’s simple, visual, and effective for moderate-sized datasets like mine. The Elbow point appeared at k = 3, which was used for the final K-Means model.

