---
title: "Assignment 3"
author: "Shahbaz"
date: "2025-10-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(dplyr)
library(tidyr)
library(e1071)
```

# The purpose of this assignment is to apply NAIVE BAYES CLASSIFICATION to predict whether a customer will accept a personal loan based on two predictors:

#Online: whether the customer uses online banking (1 = YES, 0 = NO)
#CC: whether the customer holds a bank-issued credit card (1 = YES, 0 = NO)
# Outcome Variable : Loan: whether the customer accepted the loan offer (1 = YES, 0 = NO)

# Data Preparation
```{r}
UB <- read.csv("/Users/shahbazshaikh/Downloads/UniversalBank.csv")
head(UB)

UB <- UB %>% select(Online, CCAvg, Personal.Loan) %>%
  rename(Loan = Personal.Loan)

UB$Online <- factor(UB$Online, levels = c(0, 1), labels = c("No", "Yes"))
UB$Loan <- factor(UB$Loan, levels = c(0, 1), labels = c("No", "Yes"))

UB$CC <- ifelse(UB$CCAvg > 0, "Yes", "No")
UB$CC <- factor(UB$CC, levels = c("No", "Yes"))

set.seed(1)
n <- nrow(UB)
train_index <- sample(1:n, size = 0.6*n)
train_data <- UB[train_index, ]
valid_data <- UB[-train_index, ]

cat("Training set dimensions:", dim(train_data), "\n")
cat("Validation set dimensions:", dim(valid_data), "\n")
cat("\nFirst few rows of training data:\n")
head(train_data)
```

# Part A : Pivot Table
```{r}
pivot_table <- with(train_data, table(CC, Loan, Online))

print("Pivot Table (CC x Loan x Online):")
ftable(pivot_table)
```

#Part B : Probability from Pivot Table
```{r}
count_loan_yes <- pivot_table["Yes", "Yes", "Yes"]
total_count_cc_online_yes <- sum(pivot_table["Yes", , "Yes"])

prob_b <- count_loan_yes / total_count_cc_online_yes

cat("Count of customers with CC=Yes, Online=Yes, Loan=Yes:", count_loan_yes, "\n")
cat("Total count of customers with CC=Yes, Online=Yes:", total_count_cc_online_yes, "\n")
cat("P(Loan = Yes | CC = Yes, Online = Yes) =", round(prob_b, 4), "\n")
```

#Part C : Separate Pivot Tables
```{r}
# Table 1: Loan as function of Online
table_loan_online <- table(train_data$Loan, train_data$Online)
colnames(table_loan_online) <- c("Online=No", "Online=Yes")
rownames(table_loan_online) <- c("Loan=No", "Loan=Yes")

print("Pivot Table: Loan by Online")
print(table_loan_online)

# Table 2: Loan as function of CC
table_loan_cc <- table(train_data$Loan, train_data$CC)
colnames(table_loan_cc) <- c("CC=No", "CC=Yes")
rownames(table_loan_cc) <- c("Loan=No", "Loan=Yes")

print("Pivot Table: Loan by CC")
print(table_loan_cc)
```

# Part D : Compute Required Probabilities
```{r}
# Total training observations
n_train <- nrow(train_data)

p_cc1_loan1 <- table_loan_cc["Loan=Yes", "CC=Yes"] / sum(table_loan_cc["Loan=Yes", ])
p_online1_loan1 <- table_loan_online["Loan=Yes", "Online=Yes"] / sum(table_loan_online["Loan=Yes", ])
p_loan1 <- sum(table_loan_cc["Loan=Yes", ]) / n_train
p_cc1_loan0 <- table_loan_cc["Loan=No", "CC=Yes"] / sum(table_loan_cc["Loan=No", ])
p_online1_loan0 <- table_loan_online["Loan=No", "Online=Yes"] / sum(table_loan_online["Loan=No", ])
p_loan0 <- sum(table_loan_cc["Loan=No", ]) / n_train


results_d <- data.frame(
  Probability = c("P(CC=Yes | Loan=Yes)", "P(Online=Yes | Loan=Yes)", "P(Loan=Yes)",
                  "P(CC=Yes | Loan=No)", "P(Online=Yes | Loan=No)", "P(Loan=No)"),
  Value = round(c(p_cc1_loan1, p_online1_loan1, p_loan1,
                  p_cc1_loan0, p_online1_loan0, p_loan0), 4)
)

print(results_d)
```

#Part E : Naive Bayes Probability
```{r}
# Naive Bayes calculation
numerator <- p_cc1_loan1 * p_online1_loan1 * p_loan1
denominator <- numerator + (p_cc1_loan0 * p_online1_loan0 * p_loan0)
prob_e <- numerator / denominator

cat("Naive Bayes Calculation:\n")
cat("P(Loan = Yes | CC = Yes, Online = Yes) =", round(prob_e, 4), "\n")
```

#Part F: Comparison
```{r}
cat("Comparison of Probability Estimates:\n")
cat("From Pivot Table (Part B):", round(prob_b, 4), "\n")
cat("From Naive Bayes (Part E):", round(prob_e, 4), "\n")

cat("\nWhich is more accurate?\n")
cat("The pivot table estimate is more accurate when we have sufficient data in the specific combination,\n")
cat("as it doesn't rely on the conditional independence assumption of Naive Bayes.\n")
```
#Part G: Naive Bayes Model Output
```{r}
# Train Naive Bayes model
nb_model <- naiveBayes(Loan ~ Online + CC, data = train_data)

# Examine model output
print("Naive Bayes Model Output:")
print(nb_model)

# Predict probability for our case
test_case <- data.frame(Online = "Yes", CC = "Yes")
pred_probs <- predict(nb_model, test_case, type = "raw")
prob_g <- pred_probs[, "Yes"]

cat("\nProbability from naiveBayes function:\n")
cat("P(Loan = Yes | CC = Yes, Online = Yes) =", round(prob_g, 4), "\n")

cat("\nComparison with manual calculation:\n")
cat("Manual Naive Bayes (Part E):", round(prob_e, 4), "\n")
cat("R naiveBayes function:", round(prob_g, 4), "\n")
cat("Are they equal?", abs(prob_e - prob_g) < 0.0001, "\n")
```

### Conclusion
Both the empirical (pivot table) and theoretical (Naive Bayes) estimates produced nearly identical probabilities (≈0.0965). This confirms that the independence assumption of Naive Bayes performs well for these variables, and that customers who both own a credit card and use online banking have about a 9.6% chance of accepting a personal loan.

